Revealing transient strain in geodetic data with Gaussian process regression

### 2017-08-13 RevisedSubmission/hines_hetland_2017_gji.tex
added more

\subsection{Transient Strain Rates}\label{sec:Results} 
Having established a noise model and a prior for transient displacements, we use the cleaned GNSS dataset to calculate transient strain rates in the Puget Sound region.  We calculate transient strain rates for each day from January 1, 2010 to May 15, 2017. The strain rates are estimates at a grid of points spanning the study area. In Figure \ref{fig:StrainMap} we show the transient strain rates on January 1, 2016, which coincides with the height of an SSE. We have included an animation showing the map view of strain rates through time as supplementary material. The strain rates shown in Figure \ref{fig:StrainMap} are generally similar to the strain rates during the other six SSEs considered in this study. The SSEs cause trench perpendicular compression in the Olympic Peninsula and extension east of Puget Sound. The strain transitions from compressional to extensional strain around the southern tip of Vancouver Island, which coincides with the location of where thrust slip tends to be inferred for SSEs in the Puget Sound region \citep[e.g.,][]{Dragert2001,Wech2009,Schmidt2010}. Thus, this pattern of strain is to be expected. During the period in between SSEs, secular strain rates indicate trench perpendicular compression throughout this study region \citep{Murray2000,McCaffrey2007,McCaffrey2013}. When comparing inferred strain rates from SSEs to the secular strain rates, we see that SSEs are concentrating tectonically accumulated strain energy towards the trench, and presumably pushing the subduction zone closer to failure.

@@

In Figure \ref{fig:StrainTs} we show the time dependence of estimated transient strain rates at a position on the Olympic Peninsula, where transient strain rates from SSEs are largest. To verify that the estimated transient strain rates are accurately identifying geophysical signal, we compare the signal-to-noise ratio from eq. (\ref{eq:SNR}) to the frequency of seismic tremor (Figure \ref{fig:StrainMag}). A signal-to-noise ratio greater than ${\sim}3$ can be interpreted as a detected geophysical signal. For each detected event there is a corresponding peak in seismic tremor. We are also able to clearly identify transient strain associated with a more subtle SSE in August 2014. In between peaks in seismic tremor, the signal-to-noise ratio is consistently between 0 and 2, suggesting that all the transient strain detected at this location is associated with SSEs.

@@

The results we have presented thus far indicate that we are identifying the strain that we should expect to see. There are, however, subtle features in our estimated transient strain rates which we were not expecting. For example, there is a brief period of east-west extension on the Olympic Peninsula several days prior to some of the SSEs. This feature can be seen before the summer 2012 and winter 2015-2016 SSEs in Figure \ref{fig:StrainTs} as well as in the supplementary animation. While this deformation is noteworthy, a discussion on the mechanisms causing it is outside the scope of this study.

>>>>

\subsection{Transient Strain Rates}\label{sec:Results} 
Having established a noise model and a prior for transient displacements, we use the cleaned GNSS dataset to calculate transient strain rates in the Puget Sound region.  We calculate transient strain rates for each day from January 1, 2010 to May 15, 2017. The strain rates are estimates at a grid of points spanning the study area. In Figure \ref{fig:StrainMap} we show the transient strain rates on January 1, 2016, which coincides with the height of an SSE. We have included an animation showing the map view of strain rates through time as supplementary material. The strain rates shown in Figure \ref{fig:StrainMap} are generally similar to the strain rates during the other six SSEs considered in this study. The SSEs cause trench perpendicular compression in the Olympic Peninsula and extension east of Puget Sound. The strain transitions from compression to extension around the southern tip of Vancouver Island, which coincides with the location of thrust slip for SSEs in the Puget Sound region \citep[e.g.,][]{Dragert2001,Wech2009,Schmidt2010}. Thus, this pattern of strain is to be expected. During the period in between SSEs, secular strain rates indicate trench perpendicular compression throughout this study region \citep{Murray2000,McCaffrey2007,McCaffrey2013}. When comparing inferred strain rates from SSEs to the secular strain rates, we see that SSEs are concentrating tectonically accumulated strain energy towards the trench, and presumably pushing the subduction zone closer to failure.

@@

In Figure \ref{fig:StrainTs} we show the time dependence of estimated transient strain rates at a position on the Olympic Peninsula, where transient strain rates from SSEs are largest. To verify that the estimated transient strain rates are accurately identifying geophysical signal, we compare the signal-to-noise ratio from eq. (\ref{eq:SNR}) to the frequency of seismic tremor (Figure \ref{fig:StrainMag}). A signal-to-noise ratio greater than ${\sim}3$ can be interpreted as a detected geophysical signal. We detect nine distinct events, which each correspond to peaks in seismic tremor. The smaller events detected in August 2014 and February 2017 can be considered inter-SSE events. They were not among the SSEs used to constrain the prior covariance function. In between peaks in seismic tremor, the signal-to-noise ratio is consistently between 0 and 2, suggesting that all the transient strain detected at this location is associated with SSEs and inter-SSE events.

@@

The results we have presented thus far indicate that we are identifying the strain that we should expect to see. There are, however, subtle features in our estimated transient strain rates which we were not expecting. For example, some SSEs are preceded by a brief period of east-west extension on the Olympic Peninsula. This feature can be seen several days before the summer 2012 and winter 2015-2016 SSEs in Figure \ref{fig:StrainTs} as well as in the supplementary animation. While this deformation is noteworthy, a discussion on the mechanisms causing it is outside the scope of this study.
